---
title: "Data Understanding"
output: 
  html_document:
    keep_md: true
    css: "my_style.css"
---

<html>
<style>
details > summary {
padding: 4px;
font-size: 16px;
font-weight: bold;
width: 600px;
border: none;
cursor: pointer;
}

ul {
background-color: #eeeeee;
padding: 4px;
margin: 4;
box-shadow: 1px 1px 2px #bbbbbb;
}

details > p {
background-color: #eeeeee;
padding: 4px;
margin: 0;
box-shadow: 1px 1px 2px #bbbbbb;
}
</style>
<body>


<script src="hideOutput.js"></script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)
```


Packages used to create the below figures:

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)
library(lubridate)
library(readxl)
library(DT)
library(ggcorrplot)
```


Load partial volume corrected regional tau-PET data, as downloaded from ADNI:
```{r}
tau.df <- read.csv("../../ADNI_Data/Raw_Data/UCBERKELEYAV1451_PVC_05_12_20.csv", stringsAsFactors = T)
tau.df$EXAMDATE <- as.Date(as.character(tau.df$EXAMDATE), format="%m/%d/%Y")
```


Filter tau data to contain only subjects with 2+ tau-PET scans, and omit irrelevant columns:
```{r}
tau.df <- tau.df %>%
  select(-VISCODE, -update_stamp, -HEMIWM_SUVR, -BRAAK12_SUVR,
         -BRAAK34_SUVR, -BRAAK56_SUVR, -OTHER_SUVR) %>%
  select(!matches("VOLUME")) %>%
  group_by(RID) 

colnames(tau.df) <- str_replace_all(colnames(tau.df), "_SUVR", "")
```


As shown in [Data Understanding](https://anniegbryant.github.io/DA5030_Final_Project/Pages/2_Data_Understanding.html#htmlwidget-aee61fc0845b1248fae6), the ROIs are not precisely standardized to the inferior cerebellum gray matter SUVR. I will re-standardize the ROI SUVR values here:

```{r}
tau.stand <- tau.df
for (i in 4:ncol(tau.stand)) {
  tau.stand[i] <- tau.stand[i]/ tau.df[4]
}
rm(tau.df)
```


The next step is to select brain regions *a priori*. I am going to stratify the cortical parcellations and subcortical segmentations based on Schöll et al. [(2016)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4779187/) and per UCSF's recommendations for usage of their tau-PET data. Here is the stratification across the Braak stages:

<div class="fold s">

```{r}
roi.braak <- read.csv("roi_braak_stages.csv") %>% mutate(ROI_Name = tolower(ROI_Name)) %>%
  mutate(Hemisphere = ifelse(str_detect(ROI_Name, "rh_|right"), "Right", "Left"))
datatable(roi.braak)
```

</div>

I will filter the tau-PET dataset to only include SUVR data for ROIs detailed in the above list, by first reshaping the tau-PET SUVR data from wide to long. Then, I will merge left and right hemisphere ROIs into one bilateral ROI by taking the mean SUVR.

```{r}
tau.stand.roi <- tau.stand %>%
  pivot_longer(., cols=c(-RID, -VISCODE2, -EXAMDATE), names_to="ROI_Name", values_to="SUVR") %>%
  mutate(ROI_Name=tolower(ROI_Name)) %>%
  semi_join(., roi.braak) %>%
  left_join(., roi.braak) %>%
  mutate(ROI_Name = str_replace_all(ROI_Name, "right_|left_|ctx_rh_|ctx_lh_", "")) %>%
  group_by(RID, VISCODE2, EXAMDATE, ROI_Name, Braak) %>%
  summarise(SUVR = mean(SUVR, na.rm=T))
```

Now, I will re-shape the data back to wide to be compatible with the cognitive status data shape.

<div class="fold o">
```{r}
tau.stand.roi <- tau.stand.roi %>% 
  select(-Braak) %>%
  pivot_wider(id_cols=c(RID, VISCODE2, EXAMDATE), names_from="ROI_Name",
              values_from="SUVR")

str(tau.stand.roi)
```
</div>



ADNI compiled a merged dataset containing key information from several tables, including subject demographics, selected cognitive assessment scores, and select biomarker data.
```{r}
subj.info <- read.csv("../../ADNI_Data/Raw_Data/ADNIMERGE.csv", stringsAsFactors = T, na.strings="")
```


Filter subject demographics data to contain only the following features of interest:

* `RID`: Participant roster ID, which serves as unique subject identifier 
* `VISCODE`: Visit code
* `EXAMDATE`: Date
* `AGE`: Age at visit
* `PTGENDER`: Biological sex
* `CDRSB`: CDR Sum-of-Boxes score at visit
* `DX`: Current cognitive diagnosis


```{r}
subj.info <- subj.info %>% select(RID, VISCODE, AGE, PTGENDER, CDRSB, DX)
```

I actually can't join the two datasets on the EXAMDATE feature, as these sometimes differ by one or two days depending on when the records were entered. Instead, I will join by the RID subject identifier and VISCODE, a visit code identifier.

```{r}
full.df <- inner_join(tau.stand.roi, subj.info, by=c("RID", "VISCODE2"="VISCODE"))  %>%
  filter(!is.na(CDRSB)) %>%
  group_by(RID) %>%
  mutate(n_visits = n()) %>%
  filter(n_visits>1) %>%
  select(-n_visits)
```


Click to see the structure of this merged dataset:

<div class="fold o">
```{r}
str(full.df)
```
</div>

```{r, results='asis'}
cat("\nNumber of longitudinal tau-PET scans with accompanying cognitive data: **\n",
    nrow(full.df), "**\nNumber of subjects in merged dataset: **", 
    length(unique(full.df$RID)), "**\n", "\n", sep="")
```
As it turns out, only 588 of the original 593 tau-PET scans had corresponding cognitive assessments. This leaves 576 unique PET scan datapoints for 243 subjects. 


Lastly, before I can perform outlier detection, I need to derive the longitudinal features upon which the prediction models will be built -- namely, annual change in tau-PET SUVR and annual change in CDR-Sum of Boxes score.

<div class="fold o">
```{r}
annual.changes <- full.df %>%
  ungroup() %>%
  select(-AGE, -PTGENDER, -DX, -VISCODE2) %>%
  pivot_longer(cols=c(-RID, -EXAMDATE), names_to="Metric",
               values_to="Value") %>%
  group_by(RID, Metric) %>%
  summarise(n_years = as.numeric((EXAMDATE - lag(EXAMDATE, 
                                                 default=EXAMDATE[1]))/365),
            change = Value - lag(Value, default=Value[1])) %>%
  filter(n_years > 0) %>%
  mutate(Annual_Change = change/n_years) %>%
  select(-n_years, -change) %>%
  group_by(RID, Metric) %>%
  mutate(interval_num = row_number()) %>%
  pivot_wider(., id_cols=c(RID, interval_num), names_from=Metric,
              values_from=Annual_Change)
datatable(annual.changes[1:5])
```
</div>



Now that the datasets are merged, I can perform outlier detection. Given the multivariate nature of this dataset (i.e. multiple brain regions), I will use Cook's Distance to estimate the relative influence of each data point in a simple multiple regression model.

```{r}
cooks.distance <- cooks.distance(lm(CDRSB ~ . - RID - interval_num, data=annual.changes))
plot(cooks.distance, pch=1, cex=1)
abline(h = 4*mean(cooks.distance, na.rm = T), col = "blue")
text(x = 1:length(cooks.distance)+2, 
     y=cooks.distance, 
     labels=ifelse(cooks.distance > 4 * mean(cooks.distance, na.rm=T), 
                   names(cooks.distance),""), 
     col="blue")
```
All but one data point have relatively low Cook's distance values, while data point #224 has a relatively large Cook's distance. This suggests large residuals and leverage associated with this datapoint, which could distort model fitting and accuracy. Upon further examination of this instance:

```{r}
as.data.frame(t(annual.changes[224,])) %>%
  rownames_to_column(var="Variable") %>%
  rename("Value" = "V1") %>%
  kable(., booktabs=T) %>%
  kable_styling(full_width = F)
```

This subject exhibits very large fluctuations in tau-PET SUVR values in several brain regions for this associated time interval. Given that SUVR values typically range from 0.75-2, changes of this large magnitude is surprising, and may certainly render this data point an outlier. Fortunately, the `interval_num` of 2 indicates that this is the second time interval for this subject, so omitting this interval doesn't reduce the total number of subjects in the analysis. I will remove this data point:

```{r}
annual.changes <- annual.changes[-224,]
```


I can now finish some aspects of data exploration that depended upon refining the subject cohort as well as the features. For starters, I will visualize the correlation in annual tau change between each of the ROIs measured:

<div class="fold s">

```{r}
annual.roi <- annual.changes %>% ungroup() %>% select(-RID, -interval_num, -CDRSB)
roi.cor <- cor(annual.roi)
p.mat <- cor_pmat(annual.roi)

ggcorrplot(roi.cor, hc.order = TRUE, 
           outline.col = "white", p.mat = p.mat,
           insig = "blank") %>% ggplotly(width=800, height=700)
```
</div>


As it turns out, there are primarily positive correlations in annual rate of change in tau-PET uptake. This indicates that most of the brain regions measured here tend to show similar changes in tau-PET accumulation over time -- though nuances unseen here may help discern small but present differences as they relate to CDR-Sum of Boxes scores. The largest correlation coefficient of 0.76 pertains to the superior parietal cortex and lateral occipital cortex.