---
title: "Data understanding backup"
author: "Annie Bryant"
date: "7/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages used:
```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(plotly)
library(lubridate)
library(forcats)
```

### Load tau-PET data, as downloaded from ADNI:
```{r}
og_tau <- read.csv("../../ADNI_Data/Raw_Data/UCBERKELEYAV1451_PVC_05_12_20.csv")
str(og_tau)
```


### Examine scan date distribution:

```{r}
og_tau %>%
  select(RID, EXAMDATE) %>%
  mutate(Scan_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  plot_ly(x=~Scan_Date, type="histogram") %>%
  layout(title = 'Tau-PET Scan Date Distribution',
         xaxis = list(title = 'Scan Date',
                      zeroline = TRUE),
         yaxis = list(title = 'Number of PET Scans')) 
```

### Examine number of longitudinal PET scans per subject (i.e. average number of PET scans each subject had):

```{r}
p_num_long <- og_tau %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_scans=n()) %>%
  ggplot(., aes(x=fct_reorder(RID, n_scans, .desc=T), y=n_scans)) +
  geom_bar(stat="identity", aes(fill=n_scans, color=n_scans)) +
  labs(fill="Count", color="Count") +
  ggtitle("Number of Longitudinal PET Scans per Subject") +
  ylab("Number of PET Scans") +
  xlab("Subject") +
  theme(axis.text.x=element_blank(),
        plot.title=element_text(hjust=0.5)) 

ggplotly(p_num_long)
```

### How many subjects have at least two longitudinal scans?

I am going to focus exclusively on subjects with at least two scans, to examine changes in tau accumulation over time. 
```{r}
og_tau %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_scans=n()) %>%
  filter(n_scans>=2) %>%
  ungroup() %>%
  summarise(`Number of Subjects`=n(),
            `Number of Scans in Total`=sum(n_scans)) %>%
  kable(., booktabs=T)
```

### What is the typical time interval between consecutive tau-PET scans?

The ADNI tau-PET scans are intended to be spaced approximately one year apart for each subject, but the temporal distribution should still be examined:

```{r}
p_pet_interval <- og_tau %>%
  select(RID, EXAMDATE) %>%
  mutate(Scan_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  group_by(RID) %>%
  mutate(n_scans=n()) %>%
  filter(n_scans>=2) %>%
  mutate(Years_between_Scans = 
           as.numeric((Scan_Date - lag(Scan_Date, 
                                       default = Scan_Date[1]))/365)) %>%
  filter(Years_between_Scans>0) %>%
  ggplot(., aes(x=Years_between_Scans)) +
  geom_histogram(stat="count", color="dodgerblue3") +
  ggtitle("Years in between Tau-PET Scans per Subject") +
  ylab("Frequency") +
  xlab("# Years between two consecutive scans for a subject") +
  theme_minimal() +
  theme(plot.title=element_text(hjust=0.5)) 

ggplotly(p_pet_interval)
```

### Is there any missing tau-PET SUVR uptake data for the measured ROIs?

```{r}
og_tau %>%
  select(-VISCODE, -VISCODE2, -update_stamp) %>%
  group_by(RID) %>%
  mutate(n_scans=n()) %>%
  filter(n_scans>=2) %>%
  select(-n_scans) %>%
  ungroup() %>%
  select(!matches("VOLUME")) %>%
  pivot_longer(cols=c(-RID, -EXAMDATE), names_to="ROI", values_to="SUVR") %>%
  group_by(ROI) %>%
  summarise(`Percent Missing` = sum(is.na(SUVR))/n(),
            `Number Missing` = sum(is.na(SUVR))) %>%
  kable(., booktabs=T)
```

### What is the average tau-PET SUVR uptake by ROI?

```{r}
p_roi_suvr <- og_tau %>%
  select(-VISCODE, -VISCODE2, -update_stamp) %>%
  group_by(RID) %>%
  mutate(n_scans=n()) %>%
  filter(n_scans>=2) %>%
  select(-n_scans) %>%
  select(!matches("VOLUME")) %>%
  pivot_longer(cols=c(-RID, -EXAMDATE), names_to="ROI", values_to="SUVR") %>%
  mutate(ROI = str_replace(ROI, "_SUVR", "")) %>%
  group_by(ROI) %>%
  summarise(Mean_SUVR=mean(SUVR, na.rm=T),
            SD_SUVR = sd(SUVR, na.rm=T),
            ymin = Mean_SUVR-SD_SUVR,
            ymax = Mean_SUVR+SD_SUVR) %>%
  ggplot(data=., mapping=aes(x=fct_reorder(ROI, Mean_SUVR, .desc=F), 
                             y=Mean_SUVR, fill=Mean_SUVR,
                             label = ROI)) +
  geom_bar(stat="identity", show.legend=F) +
  geom_errorbar(aes(ymin=ymin, ymax=ymax), width=0) +
  coord_flip() +
  theme_minimal() +
  ylab("Mean Tau-PET SUVR") +
  xlab("Region of Interest") +
  ggtitle("Mean Tau-PET SUVR by ROI")


ggplotly(p_roi_suvr, height=1000, width=600, tooltip=c("label", "y"))
```

### Load ADAS cognitive score data, as downloaded from ADNI:
```{r}
adas <- read.csv("../../ADNI_Data/Raw_Data/ADAS_ADNIGO23.csv")
str(adas)
```


The primary variable of interest is `TOTSCORE`, reflecting the sum of the ADAS component sub-scores.

```{r}
summary(adas$TOTSCORE)
```
As verified using `summary()`, this metric ranges from 0 (no cognitive impairment) to 70 (severe cognitive impairment). There are 61 missing data points for this feature, though there are far more records here than will actually be used for analysis. This dataset includes subjects without tau-PET images as well as exam sessions that did not coincide with tau-PET scan sessions, both of which will be excluded once the datasets are merged.


### Examine ADAS cognitive assessment date distribution:

```{r}
adas %>%
  select(RID, USERDATE) %>%
  mutate(Exam_Date = as.Date(USERDATE, format="%m/%d/%Y")) %>%
  plot_ly(x=~Exam_Date, type="histogram") %>%
  layout(title = 'ADAS-13 Cognitive Assessment Date Distribution',
         xaxis = list(title = 'Exam Date',
                      zeroline = TRUE),
         yaxis = list(title = 'Number of Assessed Subjects')) 
```


### Examine number of longitudinal ADAS-13 cognitive assessments per subject:

```{r}
p_adas_long <- adas %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_exams=n()) %>%
  ggplot(., aes(x=fct_reorder(RID, n_exams, .desc=T), y=n_exams, label=RID)) +
  geom_bar(stat="identity", aes(fill=n_exams, color=n_exams)) +
  labs(fill="Count", color="Count") +
  ggtitle("Number of Longitudinal ADAS-13 Assessments per Subject") +
  ylab("Number of ADAS-13 Scores") +
  xlab("Subject") +
  theme(axis.text.x=element_blank(),
        plot.title=element_text(hjust=0.5)) 

ggplotly(p_adas_long, tooltip = c("label", "y"))
```


### How many subjects have at least two ADAS-13 cognitive assessments?

I am going to focus exclusively on subjects with at least two longitudinal ADAS-13 assessments, to predict change over time based on regional tau accumulation. 
```{r}
adas %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  ungroup() %>%
  summarise(`Number of Subjects`=n(),
            `Number of ADAS-13 Assessments in Total`=sum(n_exams)) %>%
  kable(., booktabs=T)
```

### What is the typical time interval between consecutive ADAS-13 assessments?

While ADNI subjects are typically assessed annually, the temporal distribution should still be examined:

```{r}
p_adas_interval <- adas %>%
  select(RID, USERDATE) %>%
  mutate(Exam_Date = as.Date(USERDATE, format="%m/%d/%Y")) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  arrange(Exam_Date) %>%
  mutate(Years_between_ADAS13 = 
           as.numeric((Exam_Date - lag(Exam_Date, 
                                       default = Exam_Date[1]))/365)) %>%
  filter(Years_between_ADAS13>0) %>%
  ggplot(., aes(x=Years_between_ADAS13)) +
  geom_histogram(stat="count", color="dodgerblue3") +
  ggtitle("Years in between ADAS-13 Assessments per Subject") +
  ylab("Frequency") +
  xlab("# Years between two consecutive ADAS-13 assessments") +
  theme_minimal() +
  theme(plot.title=element_text(hjust=0.5)) 

ggplotly(p_adas_interval)
```


### What is the ADAS-13 total score distribution?

```{r}
p_adas_scores <- adas %>%
  mutate(Exam_Date = as.Date(USERDATE, format="%m/%d/%Y")) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  ggplot(data=., mapping=aes(x=TOTSCORE)) +
  geom_histogram(aes(y=..count..)) +
  theme_minimal() +
  ylab("# of Occurences") +
  xlab("ADAS-13 Total Score") +
  ggtitle("Distribution of ADAS-13 Total Scores")

ggplotly(p_adas_scores)
```

### Load general cognition data, as downloaded from ADNI:
```{r}
cog.df <- read.csv("../../ADNI_Data/Raw_Data/DXSUM_PDXCONV_ADNIALL.csv")
str(cog.df)
```
A value of -4 indicates that the value is missing.

This dataset is a bit tricky, as certain columns only pertain to specific ADNI cohorts. For my purposes, I am interested in the following columns: 

* DXCURREN: current cognitive diagnosis (ADNI1)
    1=NL;2=MCI;3=AD
    
* DXCONV: has there been a change in cognitive status from previous visit? (ADNI1)
    1=Yes - Conversion;2=Yes - Reversion; 0=No
      Conversion = CN to MCI or MCI to dementia
      Reversion = dementia to MCI or MCI to CN


* DXCHANGE: change in cognitive status from previous visit (ADNI2. ADNI-GO)
    1=Stable: NL; 2=Stable: MCI; 3=Stable: Dementia; 4=Conversion: NL to MCI; 5=Conversion: MCI to Dementia; 6=Conversion: NL to Dementia; 7=Reversion: MCI to NL; 8=Reversion: Dementia to MCI; 9=Reversion: Dementia to NL


* DIAGNOSIS: current cognitive diagnosis at date of exam (ADNI3)
  1=Yes - Conversion;2=Yes - Reversion; 0=No

```{r}
summary(cog.df %>% select(DXCURREN, DXCONV, DXCHANGE, DIAGNOSIS))
```

The large numbers of NA values in each column are to be expected given the disjointed nature of the features. These distinct features will be integrated in Part Three, Data Preparation.


### Examine general cognitive assessment date distribution:

```{r}
cog.df %>%
  select(RID, EXAMDATE) %>%
  mutate(Exam_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  plot_ly(x=~Exam_Date, type="histogram") %>%
  layout(title = 'General Cognitive Assessment Date Distribution',
         xaxis = list(title = 'Exam Date',
                      zeroline = TRUE),
         yaxis = list(title = 'Number of Assessed Subjects')) 
```

It seems that one subject's exam date was incorrectly inputted as 2109: 

```{r}
cog.df %>%
  mutate(Exam_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  mutate(Year=year(Exam_Date)) %>%
  filter(Year==2109) %>%
  select(-PTID) %>%
  select(Phase:EXAMDATE) %>%
  kable(., booktabs=T)
```

Since the data was entered into the ADNI databse in 2019, a reasonable assumption would be that the exam date should read 7/26/2019 instead of 7/26/2109. This will be addressed in Data Preparation. For now, I'll filter that data point out to visualize the temporal distribution better:

```{r}
cog.df %>%
  select(RID, EXAMDATE) %>%
  mutate(Exam_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  filter(year(Exam_Date)!=2109) %>%
  plot_ly(x=~Exam_Date, type="histogram") %>%
  layout(title = 'General Cognitive Assessment Date Distribution',
         xaxis = list(title = 'Exam Date',
                      zeroline = TRUE),
         yaxis = list(title = 'Number of Assessed Subjects')) 
```


### Examine number of longitudinal general cognitive assessments per subject:

```{r}
p_cog_long <- cog.df %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_exams=n()) %>%
  ggplot(., aes(x=fct_reorder(RID, n_exams, .desc=T), y=n_exams, label=RID)) +
  geom_bar(stat="identity", aes(fill=n_exams, color=n_exams)) +
  labs(fill="Count", color="Count") +
  ggtitle("Number of Longitudinal General Cognitive Assessments per Subject") +
  ylab("Number of Cognitive Assessments") +
  xlab("Subject") +
  theme(axis.text.x=element_blank(),
        plot.title=element_text(hjust=0.5)) 

ggplotly(p_cog_long, tooltip = c("label", "y"))
```

### How many subjects have at least two general cognitive assessments?

I am going to focus exclusively on subjects with at least two longitudinal general cognitive assessments, to predict conversion to a more severely impaired cognitive status based on regional tau accumulation over time. 
```{r}
cog.df %>%
  mutate(RID=as.character(RID)) %>%
  group_by(RID) %>%
  summarise(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  ungroup() %>%
  summarise(`Number of Subjects`=n(),
            `Number of General Cognitive Assessments in Total`=sum(n_exams)) %>%
  kable(., booktabs=T)
```

### What is the typical time interval between consecutive general cognitive assessments?

While ADNI subjects are typically assessed annually, the temporal distribution should still be examined. Note: I have filtered out the subject with the incorrectly-inputted exam date for visualization.

```{r}
p_cog_interval <- cog.df %>%
  select(RID, EXAMDATE) %>%
  mutate(Exam_Date = as.Date(EXAMDATE, format="%m/%d/%Y")) %>%
  filter(year(Exam_Date) != 2109) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  arrange(Exam_Date) %>%
  mutate(Years_between_Cog_Assess = 
           as.numeric((Exam_Date - lag(Exam_Date, 
                                       default = Exam_Date[1]))/365)) %>%
  filter(Years_between_Cog_Assess>0) %>%
  ggplot(., aes(x=Years_between_Cog_Assess)) +
  geom_histogram(stat="count", color="dodgerblue3") +
  ggtitle("Years in between General Cognitive Assessments per Subject") +
  ylab("Frequency") +
  xlab("# Years between two consecutive assessments") +
  theme_minimal() +
  theme(plot.title=element_text(hjust=0.5)) 

ggplotly(p_cog_interval)
```


### How is cognitive status distributed in the distinct ADNI cohorts?


```{r}
p_adni1_curr <- cog.df %>%
  filter(!is.na(DXCURREN)) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  mutate(DXCURREN = factor(DXCURREN, levels=c(1,2,3),
                           labels=c("Cognitively Normal",
                                    "Mild Cognitive Impairment", 
                                    "Dementia"))) %>%
  ggplot(data=., mapping=aes(x=DXCURREN, fill=DXCURREN)) +
  geom_histogram(stat="count", show.legend=F) +
  theme_minimal() +
  ylab("# of Occurences") +
  xlab("Current Diagnosis in ADNI1") +
  ggtitle("Distribution of Current Diagnosis in ADNI1") +
  theme(plot.title=element_text(hjust=0.5),
        legend.position='none')

ggplotly(p_adni1_curr)
```

```{r}
p_adni3_curr <- cog.df %>%
  filter(!is.na(DIAGNOSIS)) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  mutate(DIAGNOSIS = factor(DIAGNOSIS, levels=c(1,2,3),
                           labels=c("Cognitively Normal",
                                    "Mild Cognitive Impairment", 
                                    "Dementia"))) %>%
  ggplot(data=., mapping=aes(x=DIAGNOSIS, fill=DIAGNOSIS)) +
  geom_histogram(stat="count", show.legend=F) +
  theme_minimal() +
  ylab("# of Occurences") +
  xlab("Current Diagnosis in ADNI3") +
  ggtitle("Distribution of Current Diagnosis in ADNI3") +
  theme(plot.title=element_text(hjust=0.5),
        legend.position='none')

ggplotly(p_adni3_curr)
```


```{r}
p_adni1_change <- cog.df %>%
  filter(!is.na(DXCONV)) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  mutate(DXCONV = factor(DXCONV, levels=c(0,1,2),
                           labels=c("Stable",
                                    "Conversion", 
                                    "Reversion"))) %>%
  ggplot(data=., mapping=aes(x=DXCONV, fill=DXCONV)) +
  geom_histogram(stat="count", show.legend=F) +
  theme_minimal() +
  ylab("# of Occurences") +
  xlab("Cognitive Status Change in ADNI1") +
  ggtitle("Distribution of Cognitive Status Changes in ADNI1") +
  theme(plot.title=element_text(hjust=0.5),
        legend.position='none')

ggplotly(p_adni1_change)
```

```{r}
p_adni2_change <- cog.df %>%
  filter(!is.na(DXCHANGE)) %>%
  group_by(RID) %>%
  mutate(n_exams=n()) %>%
  filter(n_exams>=2) %>%
  mutate(DXCHANGE = factor(DXCHANGE, levels=c(1:9),
                           labels=c("Stable CN",
                                    "Stable MCI",
                                    "Stable Dementia",
                                    "CN to MCI",
                                    "MCI to Dementia",
                                    "CN to Dementia",
                                    "MCI to CN",
                                    "Dementia to MCI",
                                    "Dementia to CN"))) %>%
  ggplot(data=., mapping=aes(x=DXCHANGE, fill=DXCHANGE)) +
  geom_histogram(stat="count", show.legend=F) +
  theme_minimal() +
  ylab("# of Occurences") +
  xlab("Cognitive Status Change in ADNI2") +
  ggtitle("Distribution of Cognitive Status Changes in ADNI2") +
  theme(plot.title=element_text(hjust=0.5),
        legend.position='none',
        axis.text.x=element_text(angle=45))

ggplotly(p_adni2_change)
```